---
title: "Condition-comparable GRNs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Condition-comparable GRNs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Goal

This workflow estimates a GRN for each condition of the same cell type while
keeping coefficient differences in a common model space. The default
`shared_design_independent` engine deliberately separates two concerns:

1. Comparability is enforced by one complete edge dictionary, pooled
   final-predictor scaling, and, within each target, one regularization path
   and selected regularization level shared by conditions.
2. Biological condition coefficients are estimated independently and are not
   shrunk toward each other by a cross-condition group penalty.

This is the closest shared-design analogue of running separate Pando models,
without allowing edge definitions or coefficient units to drift.

## Fit the shared design

Start with the same normalized paired RNA/ATAC `GRNData` used by the standard
Pando workflow, including motif matches.

```{r fit}
condition_object <- infer_condition_grn(
  grn_object,
  cell_type_col = "cell_type",
  condition_col = "condition",
  genes = metabolic_genes,
  network_name = "condition_grn",
  method = "shared_design_independent",
  candidate_screen = "condition_union",
  alpha = 0.5,
  condition_mix = 1,
  condition_weight = "equal",
  reference_condition = "Control",
  nfolds = 5,
  lambda_selection = "lambda.1se",
  scale = TRUE,
  seed = 12345
)
```

For `candidate_screen = "condition_union"`, an edge is eligible in a condition
only when both its TF and its motif-bearing peak pass screening in that same
condition. Pando takes the union of those complete edges across conditions; it
does not independently union TFs and peaks before pairing them.

With `scale = TRUE`, Pando first forms each raw interaction

\[
x_e = \mathrm{RNA}_{TF(e)} \times \mathrm{ATAC}_{peak(e)}
\]

and then stores one pooled center and scale for that complete edge predictor
across the cell type. The target is standardized once over the same pooled
cells. Every condition therefore uses identical coefficient units.

## Inspect the lossless fit contract

```{r inspect}
fit <- condition_grn_fit(
  condition_object,
  network_name = "condition_grn",
  cell_type = "Tumor_cell"
)

fit$schema_version
fit$fit_engine
fit$edge_table
fit$eligibility_mask
fit$beta
fit$contrast
fit$predictor_transform
fit$response_transform
fit$target_fit
fit$target_rsq
```

The explicit contrast is

\[
\Delta\beta_{e,c} = \beta_{e,c} - \beta_{e,reference}.
\]

The reference column is therefore exactly zero. The `eligibility_mask` should
always accompany the coefficient matrix so that a structural absence is not
mistaken for a fitted zero.

## Downstream model-space projection

A downstream method can reproduce Pando's model space for edge \(e\) in unit
\(u\) with the stored transform:

\[
z_{e,u} =
\frac{
  \mathrm{RNA}_{TF(e),u}\mathrm{ATAC}_{peak(e),u}
  - \mathrm{center}_e
}{
  \mathrm{scale}_e
}.
\]

Condition differences should be projected as
\(\sum_e \Delta\beta_{e,c}z_{e,u}\). Do not renormalize
\(\Delta\beta\) by its absolute sum: that would erase the fitted effect
magnitude that the shared scale was designed to preserve.

## Compatibility networks

```{r compatibility}
Params(condition_object)$condition_network_index
coef(
  condition_object,
  network = "condition_grn__Tumor_cell__condition__Drug"
)
```

These remain standard Pando `Network` objects. The Universal network stores an
equal-condition coefficient mean for existing visualization interfaces only;
it is not used as the baseline for condition contrasts.
