---
title: "Condition-specific sub-GRNs and RegCompass handoff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Condition-specific sub-GRNs and RegCompass handoff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Two Pando workflows

Pando retains its original `infer_grn()` workflow unchanged. Use
`infer_condition_grn()` when conditions of the same cell type must be compared
directly. The condition workflow still writes standard Pando `Network` objects,
so `coef()`, `gof()` and `NetworkParams()` continue to work. Its
`ConditionGRNFit` and projection objects add the lossless information required
by RegCompass.

## Fit one shared candidate graph

```{r fit}
condition_object <- infer_condition_grn(
  grn_object,
  cell_type_col = "cell_type",
  condition_col = "condition",
  genes = metabolic_genes,
  network_name = "condition_grn",
  candidate_screen = "pooled_within_condition",
  alpha = 0.5,
  condition_mix = 0.5,
  condition_weight = "equal",
  reference_condition = "Control",
  nfolds = 5,
  lambda_selection = "lambda.1se",
  scale = TRUE,
  seed = 12345
)
```

Pando forms each candidate interaction on paired single cells:

\[
x_{i,e}=\mathrm{RNA}_{i,TF(e)}\mathrm{ATAC}_{i,peak(e)}.
\]

Every condition uses the same motif/domain edge dictionary and the same pooled
edge and target transforms. `pooled_within_condition` scores each candidate by
the weighted root-mean-square of its within-condition correlations. This
removes pure between-condition mean shifts without cancelling a positive effect
in one condition against a negative effect in another.

Sparse multi-task selection may choose different active edges in each
condition. The common-scale refit keeps an inactive but estimable edge at exact
zero and an unavailable edge at `NA`.

## Standard Pando-compatible networks

```{r networks}
Params(condition_object)$condition_network_index

control_network <- GetNetwork(
  condition_object,
  network = "condition_grn__Tumor_cell__condition__Control"
)
coef(control_network)
gof(control_network)
NetworkParams(control_network)
```

The shared network is
`condition_grn__<cell_type>__shared`; condition networks are
`condition_grn__<cell_type>__condition__<condition>`. Their coefficient tables
retain the original Pando columns:

```text
tf  target  region  term  estimate  corr
```

`estimate` is the absolute refitted condition coefficient, not a reference
contrast. An unavailable edge has `estimate = NA`; an estimable inactive edge
has `estimate = 0`.

## Lossless v3 contract

```{r contract}
fit <- condition_grn_fit(
  condition_object,
  network_name = "condition_grn",
  cell_type = "Tumor_cell"
)

fit$beta_condition_std
fit$beta_shared_std
fit$delta_condition_std
fit$active_mask
fit$estimability_mask
fit$absolute_direction
```

For every shared candidate edge:

\[
\beta_{e,c}=\beta_{e,shared}+\delta_{e,c}.
\]

Different condition sub-GRNs are induced by their own active masks, so their TF,
peak and target node sets may differ. Effects may also reverse sign. These
differences do not change the edge coordinate system or coefficient units.

```{r compare}
condition_grn_subgraph(fit, "Control")
condition_grn_subgraph(fit, "Drug")

contrast <- condition_grn_contrast(fit, "Control", "Drug")
contrast[, c(
  "tf", "region", "target",
  "beta_condition_1", "beta_condition_2",
  "delta_beta", "pairwise_direction", "sign_switch", "comparable"
)]
```

`delta_condition` means deviation from the shared coefficient. It is not the
same as repression: absolute regulatory direction is always
`sign(beta_condition)`.

## Project before constructing metacells

```{r projection}
projection <- project_condition_grn_cells(
  condition_object,
  fit = fit,
  component = "condition",
  scale = "std",
  output = "gene_score",
  targets = metabolic_genes,
  nonestimable = "propagate"
)

projection$gene_score
projection$gene_direction
projection$target_condition_status
projection$aggregation_contract
```

For a cell in condition \(c\), Pando computes

\[
R_{i,g,c}=\sum_{e\rightarrow g}z_{i,e}\beta_{e,c}
\]

with the exact transforms stored during fitting. The score is signed. If any
candidate edge needed by a target is unavailable in that condition, the target
score is `NA`; it is never silently converted to zero.
The projection also verifies the exact fitted cell IDs, RNA/ATAC assays and
condition labels before calculating scores.

RegCompass should construct metacells only after this projection and aggregate
each target column by the arithmetic mean within
cell type × condition × sample/donor. It must not multiply aggregated TF and
peak values, recompute TF-IDF or pooled transforms, refit Pando coefficients, or
replace `NA` with zero. Because every condition projection has identical target
columns and scale, metacells from different conditions remain directly
comparable for the same GEM reaction.
